\documentclass{article}
\usepackage[submission]{colm2026_conference}

\usepackage{microtype}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{lineno}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,calc}

\input{math_commands.tex}

\definecolor{boxblue}{RGB}{0,114,178}
\definecolor{pastein}{RGB}{213,94,0}
\definecolor{timegray}{RGB}{80,80,80}
\definecolor{darkblue}{rgb}{0, 0, 0.5}
\hypersetup{colorlinks=true, citecolor=darkblue, linkcolor=darkblue, urlcolor=darkblue}

\title{Don't Paste In Conditioning Tokens: Noising Conditioning Tokens \\
for Conditional Generation in Continuous Diffusion Language Models}

% Authors must not appear in the submitted version (handled by the "submission" option).
\author{Anonymous}

\begin{document}

\ifcolmsubmission
\linenumbers
\fi

\maketitle

\begin{abstract}
Fundamentally, diffusion and flow models learn a score or velocity which can be integrated
via an ODE solver to map from a base distribution to data distribution. One way to
parameterize their learning objective is via the $x_1$ prediction; i.e, predict the target
data given the current noised input $x_t$. In practical setting for language, generation is
conditional: given a prompt, generate a response. Motivated by the downstream application of
conditional generation, we view the $x_1$ denoising objective of continuous diffusion language
models as conditional prediction: given an input $x_t$ (and optionally conditioning info $c$),
predict the target output data $x_1$. Under this view, we see the various noising levels of
the training objective as curriculum learning; predict the data $x_1$ varying degrees of
signal of the input $x_t$. From this perspective, we introduce a simple modification to
continuous diffusion language model training objective: noise the conditioning tokens as well.
We demonstrate that with this simple modification, we get better generalization and generation
quality, as measured by various datasets such as Sudoku, NQueens, and moreover (TODO ADD)
language reasoning with GSM8k. Our method can be seen as a generalization of the classifier
free guidance training set up.


% TODO: ~150 words. Core claim: standard conditional continuous-diffusion LMs hold
% conditioning (clue) tokens clean throughout corruption ("paste-in"). We instead noise
% the conditioning tokens during training, controlled by p_condclean (prob a conditioning
% token is kept clean). Lower p_condclean (more noising) consistently improves infill
% accuracy and, especially, solution COVERAGE on Sudoku (easy) and N-Queens (8x8, 10x10).
% Headline numbers: Sudoku-easy N=10k acc 86.6% (p=0.2) vs 56.05% (p=1.0);
% N-Queens coverage monotonic across all settings. Position as generalization of CFG.
\end{abstract}

\section{Introduction}
\label{sec:intro}
% TODO: Motivate conditional generation in continuous diffusion LMs (infilling puzzles:
%   observe clue cells, generate the rest). Standard practice: clamp conditioning tokens
%   clean at every denoising step ("paste-in" / always-clean clues, our p_condclean=1.0).
% TODO: State the surprising finding: NOISING conditioning tokens during training improves
%   conditional generation; clean-clue training collapses toward fewer modes.
% TODO: Frame as a generalization of classifier-free guidance (p=1 conditional-only,
%   p=0 unconditional-only). See notes June 11-12.
% TODO: Contributions bullet list (the noised-conditioning training scheme; cross-task
%   evidence on Sudoku + N-Queens; coverage as the discriminating metric).
% Citation needs: \cite{TODO-CFG}, \cite{TODO-FLM}, \cite{TODO-diffusion-forcing}.


Fundamentally, diffusion and flow models learn a score or velocity corresponding to an ODE
which can be integrated to map from a base distribution to the target data distribution.
(we need to make this more precise) Their objective can be parameterized in different ways,
including the denoising parametrization $\hat{x}_1(x_t)$, where the goal is to predict a
clean target data $x_1$ given a current noisy input $x_t$.

Our framework is to view this learning objective from the lens of conditional prediction:
given an noisy input, predict a clean target. We then see the varying levels of noise
corruption of the input as a form of curriculum learning, where the model is trained on
harder (more noisier) inputs and easier (less noiser) inputs.

In practice for language modelling, conditional generation is the main paradigm: given a
conditioning prompt, generate a response. Currently, conditional generation in continuous
diffusion language models / flow language models utilize following learning objective: given
conditioning tokens $x^c$ and a noised response $y_t$, predict the target response $y_1$.
Likewise in sampling, the conditioning tokens are fixed and non-noised. (Double check: but
this is what sFLM does. Check FLM. Check ELF.) We improve this conditional learning objective
by allowing conditional tokens to also be noised: given noised conditional input $x_t^c$,
noised response $y_t$, predict the target response $y_1$. We believe this objective induces
curriculum learning, where the prediction problem $x \to y$ is made more difficult by
increasing the amount of noise in the conditioning input. Moreover, (TODO:  we can delete
this) by noising the conditioning input, our model has a wider target that it can learn from
(also needs to match the denoised conditioning input), making learning more compute
efficient. Finally, by varying the amount of conditioning signal, we believe that the
learning objective is more robust and conducive to generalization. Varying the level of noise
on the conditioning input can be seen as a generalization of classifier free guidance.
(TODO: verify CFG).
Empirically, we verify on various reasoning tasks, such as Sudoku and NQueens. We find that
by additionally noising the conditioning inputs, we get better conditional generation
performance on these reasoning benchmarks and we get higher levels of generation diversity.


\section{Background and Related Work}
\label{sec:background}
\paragraph{Continuous diffusion and flow language models.}
Continuous diffusion and flow models for text learn a velocity or score that an ODE/SDE solver
integrates from a base distribution to data; the denoising ($x_1$-prediction) parameterization
predicts the clean target from a noised input~\citep{TODO-FLM}. Conditional generation is typically
implemented by holding the conditioning tokens clean throughout corruption and re-pasting them at
every sampling step. Diffusion forcing~\citep{TODO-diffusion-forcing} assigns each token its own
noise level, which is what lets us noise clue tokens independently of the cells we generate.

\paragraph{Classifier-free guidance.}
Classifier-free guidance~\citep{TODO-CFG} jointly trains a conditional and an unconditional model by
randomly dropping the conditioning signal. Our scheme generalizes this: rather than a binary
present/absent conditioning signal, the per-token noise level continuously interpolates the strength
of the conditioning, with $p=1$ (always clean) and $p=0$ (always noised) as the conditional-only and
unconditional-only endpoints.

\paragraph{Corrupting conditioning / prefix tokens.}
Corrupting conditioning information has appeared in other settings but with different aims. Cascaded
diffusion models~\citep{TODO-cascaded} corrupt the low-resolution conditioning image for robustness
to upstream artifacts. In diffusion language models, Blockwise SFT~\citep{TODO-blockwise-sft} reports
that corrupting the prefix \emph{hurts} autoregressive-style decoding---a contrast with our finding
that noising clue tokens helps conditional infilling---and prompt-infilling work~\citep{TODO-prompt-infill}
fine-tunes masked diffusion models for infilling without noising the conditioning. Unlike the image
cascade setting, our inference-time conditioning input is always clean; the noising acts purely as a
training-time regularizer/curriculum.

\paragraph{Combinatorial reasoning benchmarks.}
We evaluate on Sudoku~\citep{TODO-trm} and N-Queens~\citep{TODO-gram}, constraint-satisfaction tasks
where solving unseen instances requires learning the rules rather than memorizing input/output pairs;
N-Queens additionally admits many solutions per board, enabling a coverage (diversity) metric.

\section{Method}
\label{sec:method}

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  >={Stealth[length=2.4mm]},
  font=\small,
]
\def\S{6}
% ---- our training region: the full box ----
\fill[boxblue!10] (0,0) rectangle (\S,\S);
\draw[boxblue!55,line width=0.8pt] (0,0) rectangle (\S,\S);
% ---- axes ----
\draw[->,thick] (-0.2,0) -- (\S+1.1,0) node[right]{$t$ \,(response)};
\draw[->,thick] (0,-0.2) -- (0,\S+0.7) node[above]{$t'$ \,(conditioning prompt)};
% conditioning-axis endpoints (noise/clean)
\node[left] at (-0.1,0) {\scriptsize noise};
\node[left] at (-0.1,\S) {\scriptsize clean};
% response-axis endpoints (t=0 / t=1 ticks)
\node[below] at (0,-0.18) {\scriptsize $t{=}0$};
\node[below] at (\S,-0.18) {\scriptsize $t{=}1$};
% ---- paste-in: the top edge t'=1 ----
\draw[pastein,line width=2.2pt] (0,\S) -- (\S,\S);
\node[pastein,above=2pt,align=center] at (\S/2,\S)
  {\textbf{paste-in (default):} conditioning fixed clean, $t'{=}1$};
\node[pastein,below=2pt,align=center] at (\S*0.30,\S)
  {\scriptsize limited training coverage};
% ---- ours label inside the box ----
\node[boxblue!85!black,align=center] at (\S*0.40,\S*0.46)
  {\textbf{ours:} train with various levels\\[1pt]
  of conditioning information};
% ---- expansive vertical double-arrow ----
\draw[boxblue!80!black,line width=1.3pt,<->] (\S*0.88,0.5) -- (\S*0.88,\S-0.7);
\node[boxblue!80!black,rotate=90,anchor=south] at (\S*0.88-0.18,\S*0.42)
  {\scriptsize range of conditioning noise levels};
% ---- bottom time arrow ----
\draw[timegray,line width=1.2pt,->] (0.15,-1.0) -- (\S-0.05,-1.0);
\node[timegray,below=1pt] at (\S/2,-1.0)
  {\scriptsize generation: \ noise $(t{=}0)\ \to\ $clean $(t{=}1)$};
\end{tikzpicture}
\caption{Our method. Standard conditional training pastes in clean conditioning tokens, covering
only the top edge ($t'{=}1$) of the training space (orange). Instead, we train over a range of
conditioning noise levels (blue region), exposing the model to a curriculum of conditioning
strengths. At generation time we still integrate the response from noise ($t{=}0$) to clean
($t{=}1$) with the conditioning held clean.}
\label{fig:method}
\end{figure}

% TODO: Setup. Sequence layout: in-place infill, board of length L (Sudoku 81; N-Queens n*n).
%   conditioning_mask marks clue cells; valid_tokens marks loss region (infill_loss_region=board).
% TODO: Flow / corruption process; per-token noise time tau_t under diffusion forcing.
% TODO: The knob. conditioning_time_random=true: conditioning tokens' noise time is sampled;
%   with prob p_condclean they are held clean (tau=0), else noised to the board tokens' level.
%   p=1.0 recovers standard paste-in conditional training; p=0 = fully unconditional.
% TODO: Equation for the training loss / objective.
% TODO: Sampling: Euler ODE / ancestral; conditioning positions clamped to clean one-hot at
%   inference regardless of training p (note: at inference conditioning input is always clean).
% TODO: Define metrics -- accuracy (valid-solution rate) and coverage (distinct valid
%   solutions recovered per puzzle over k samples).

[Take 1]
Our method is simple and requires minimal modification from the default
training objective of flow language models. It can also be seen as a generalization of classifier
free guidance.
Simply put, with some probability $p_{clean}$ , we keep the conditioning tokens clean; else,
we noise the conditioning tokens as well.

During sampling, following the diffusion/flow language model literature, we keep the
conditioning tokens clean and paste clean tokens at each sampling step. To prevent a large
training and sampling mismatch, with some probability $p_{clean}$ we choose to keep the
conditioning tokens clean to supply as input. The model then predicts using inputs
$(x_{t=1},y_t)$, where $x_{t=1}$ is the clean conditioning tokens and $y_t$ is the noised
response tokens.

At the same time, we do not want the model to become too reliant on the conditioning tokens and learn 
a brittle mapping $x \to y$. Thus, with the remaining
probability $1- p_{clean}$, we noise our conditioning tokens, reducing the amount of conditioning signal 
the model has access to. It receives as input
both noised conditioning tokens and noised response tokens $(x_t, y_t)$. As visualized in Figure 1, 
(TODO: add Fig link) by additionally noising our conditioning tokens, our model is trained under a
broader curriculum of conditional prediction tasks, where the model receives both a lot of 
conditioning signal (making prediction task easier) and little conditioning signal
(making prediction task harder).  

By default we also make use of the conditional tokens in the loss and have the model 
predict denoised versions of the conditioning tokens $x_t$. This utilizes all the tokens in the denoising
objective, making the training objective more compute efficient. We discuss the effect of
this in (TODO: add ablation section)

We can see our method as a generalization of CFG: we see full noise $t=0$ as analogous to
dropping the conditioning information $x_t = \varnothing$, and we see no noise $t=1$ as
preserving the conditioning tokens.

[Take 2]
TODO: Can add if it's better.

\subsection{Setup}
\label{sec:method-setup}
We work in an \emph{in-place infill} layout. Each example is a single board of length $L$
(Sudoku: $L=81$; N-Queens: $L=n^2$) over a small vocabulary. A binary
\emph{conditioning mask} marks the clue cells $\mathcal{C}$ that are revealed to the model;
the complementary generation cells $\mathcal{G}$ are to be predicted. We compute the loss over
the full board (clue and generation cells alike); only the clue cells are eligible to be held
clean (Section~\ref{sec:method-objective}).

\subsection{Corruption process and the conditioning knob}
\label{sec:method-objective}
We adopt a continuous-time flow / $x_1$-prediction model with per-token noise levels
(\emph{diffusion forcing}~\citep{TODO-diffusion-forcing}): each position $i$ carries its own time
$\tau_i \in [0,1]$, where $\tau_i = 1$ denotes the clean token and $\tau_i = 0$ full noise, and
$x_i(\tau_i)$ is the corresponding corrupted input. The denoiser $f_\theta$ receives the noised
board together with the per-token times and predicts the clean target $\hat{x}_1$.

A single scalar $p \equiv p_\text{condclean}$ controls how the clue tokens are corrupted. For each
example we draw a shared time $t \sim \mathcal{U}(0,1)$ and set
\[
\tau_i \;=\; t \quad (i \in \mathcal{G}), \qquad
\tau_i \;=\;
\begin{cases}
1 & \text{w.p. } p \quad(\text{clue held clean}),\\
t & \text{w.p. } 1-p \quad(\text{clue noised to the board level}),
\end{cases}
\quad (i \in \mathcal{C}).
\]
The training objective is the cross-entropy between the prediction and the clean board over the
loss region $\mathcal{L}$ (the full board by default):
\[
\mathcal{L}(\theta) \;=\; \mathbb{E}_{x_1,\, t,\, \{\tau_i\}}
\Big[\, \textstyle\sum_{i \in \mathcal{L}} \mathrm{CE}\big(f_\theta(x(\tau))_i,\; (x_1)_i\big) \Big].
\]
Setting $p=1$ recovers the standard ``paste-in'' conditional objective (clues always clean);
$p=0$ recovers a fully unconditional objective (clues noised exactly like the rest of the board).
Intermediate $p$ interpolates between the two, exposing the model to a curriculum of conditioning
strengths within a single training run.

\subsection{Sampling}
\label{sec:method-sampling}
At inference we follow standard practice for conditional flow language models: the conditioning
input is always clean, so clue positions are clamped to their clean one-hot values and re-pasted
after every solver step, \emph{independent of the training $p$}. We integrate the generation cells
with a first-order Euler ancestral sampler. Because training with $p<1$ exposes the model to clean
clues only a $p$ fraction of the time, a small amount of clean exposure is still required to avoid
a train/inference mismatch at the always-clean test condition; we return to this in
Section~\ref{sec:analysis}.

\subsection{Metrics}
\label{sec:method-metrics}
We report two metrics. \textbf{Accuracy} is the fraction of generated boards that are valid
solutions to the puzzle (pooled over all puzzles and samples). \textbf{Coverage} (N-Queens) is, per
puzzle, the fraction of \emph{distinct} valid solutions recovered across the $k$ samples, averaged
over puzzles; it measures the diversity of the generative distribution rather than single-shot
correctness. We report $95\%$ confidence intervals obtained by bootstrap resampling over the
evaluation puzzles ($10{,}000$ resamples; for N-Queens the resampling is over the $750$ input
boards, with per-board accuracy and coverage computed from that board's $20$ generations).

\section{Experiments}
\label{sec:experiments}

\subsection{Setup}
\label{sec:setup}
Across all experiments the backbone is a DiT-style transformer (DDiT) with hidden size $512$,
$8$ blocks, $8$ attention heads, conditioning dimension $128$, and dropout $0.1$, trained with a
flow objective and $x_1$-prediction under a log-linear noise schedule. We optimize with
Adam\footnote{The optimizer class (Adam vs.\ AdamW) is \textbf{[TODO]} to confirm against the
original configuration.} (learning rate $3\times10^{-4}$, $(\beta_1,\beta_2)=(0.9,0.999)$, no weight
decay), a constant schedule with $2{,}500$ warmup steps, gradient clipping at $1.0$,
bf16-mixed precision, and an EMA of $0.9999$. We use antithetic time sampling and a $t$-curriculum
that anneals the sampled time range from $[0,0]$ to $[0,1]$ over $100$k steps. Unless noted, we
sample with a $128$-step Euler ancestral solver at temperature $1.0$. The single swept variable is
$p_\text{condclean} \in \{0.2, 0.5, 1.0\}$ (and $0.0$ where noted); all other settings are held
fixed. Full hyperparameters are in Appendix~\ref{app:hyperparams}.
The training GPU model is \textbf{[TODO]}, and whether the vocabulary includes additional special
tokens is \textbf{[TODO]}.

\subsection{Sudoku (Easy)}
\label{sec:sudoku}
% TODO: Infill, full 9x9 board (length 81), difficulty easy, loss over board tokens.
%   Sweep p_condclean in {0.2, 0.5, 1.0}; data scales N=10,000 and N=1,000. 160k steps.
%   Eval on 2,000 puzzles. (experiment_log.md Runs 1-2; June 16 entry.)
% TODO: Table -- Sudoku-easy accuracy vs p_condclean:
%   N=10k: 0.2 -> 86.6%, 0.5 -> 74.75%, 1.0 -> 56.05%  (0.0 -> 2.85%, optional)
%   N=1k : 0.2 -> 11.0%, 0.5 -> 9.9%, 1.0 -> 2.3%       (0.0 -> 0.5%, optional)
% TODO: Observation -- monotonic: more noising (lower p) -> higher accuracy; ordering
%   preserved across data scale despite large absolute drop at N=1k.

Sudoku is a (add short introduction, can borrow from SFLM or other papers).
We believe that Sudoku is a good reasoning benchmark because of its combinatorial nature: in
order to solve unseen puzzles, the model simply learn a mapping from $x \to y$, it must
implicitly learn the constraints of the game.

We train on the \emph{easy} split in the in-place infill layout: a full $9\times9$ board
(length $81$), with the loss taken over all board tokens. We sweep $p_\text{condclean}$ at two
training-set sizes, $N=10{,}000$ and $N=1{,}000$ puzzles, training for $160$k steps at batch size
$32$ and evaluating accuracy on $2{,}000$ held-out puzzles (Table~\ref{tab:sudoku}).

\begin{table}[t]
\centering
\caption{Sudoku (easy) infill accuracy (\%) on $2{,}000$ held-out puzzles vs.\ the conditioning
clean probability $p_\text{condclean}$, with $95\%$ bootstrap CIs ($10{,}000$ resamples) in
brackets. Lower $p$ (more heavily noised clues) yields higher accuracy; the ordering is preserved
across data scale. $p{=}1.0$ is the standard ``paste-in'' setup; $p{=}0.0$ is fully unconditional.}
\label{tab:sudoku}
\begin{tabular}{lcc}
\toprule
$p_\text{condclean}$ & $N=10{,}000$ & $N=1{,}000$ \\
\midrule
$0.0$ (unconditional) & $2.85$ & $0.50$ \\
 & {\scriptsize $[2.15,\,3.60]$} & {\scriptsize $[0.20,\,0.85]$} \\
$0.2$ & $\mathbf{86.60}$ & $\mathbf{11.00}$ \\
 & {\scriptsize $[85.10,\,88.10]$} & {\scriptsize $[9.65,\,12.45]$} \\
$0.5$ & $74.75$ & $9.90$ \\
 & {\scriptsize $[72.85,\,76.65]$} & {\scriptsize $[8.60,\,11.25]$} \\
$1.0$ (paste-in) & $56.05$ & $2.30$ \\
 & {\scriptsize $[53.90,\,58.20]$} & {\scriptsize $[1.65,\,2.95]$} \\
\bottomrule
\end{tabular}
\end{table}

Accuracy decreases monotonically as $p_\text{condclean}$ increases from $0.2$ to $1.0$: at
$N=10{,}000$ the heavily-noised setting reaches $86.6\%$ versus $56.05\%$ for the standard always-clean
clues, a $30.6$-point gap, and the same ordering holds at $N=1{,}000$ despite the much lower absolute
accuracy. The fully-unconditional endpoint ($p=0.0$) collapses ($2.85\%$ / $0.5\%$): noising the
clues helps, but the model still needs to see clean clues some of the time to match the always-clean
inference condition. The best setting is thus an intermediate-but-low $p$, not the extreme. The
$95\%$ bootstrap CIs are non-overlapping across $p$ at $N=10{,}000$, so those separations are
significant; at $N=1{,}000$ the $p{=}0.2$ and $p{=}0.5$ intervals overlap (that gap is not
significant), though both clearly exceed $p{=}1.0$.

\subsection{N-Queens}
\label{sec:nqueens}
% TODO: Infill on n*n board, tokens {0=pad,1=empty,2=queen}; sweep p_condclean {0.2,0.5,1.0}.
%   Last checkpoint, 750 puzzles x 20 samples. N=8 (length 64) and N=10 (length 100).
%   (experiment_log.md Runs 3-4; June 18-19 entries.)
% TODO: Table -- accuracy & coverage vs p_condclean:
%   N=8  (750x20): 0.2 -> 99.24/97.31; 0.5 -> 98.75/91.41; 1.0 -> 97.67/88.07
%   N=10 (750x20): 0.2 -> 94.66/80.01; 0.5 -> 57.42/71.05; 1.0 -> 64.56/64.03
% TODO: Observation -- coverage monotonic across BOTH n; p=0.2 dominates. Flag the N=10
%   ACCURACY non-monotonicity (p=0.5 below p=1.0): robust claim is coverage ordering +
%   dominance of heavy noising, not strict accuracy monotonicity. Possibly single-seed noise
%   -- note variance caveat (notes Run 4).

NQueens is a (add short introduction, can borrow from GRAM.)
In additional to being combinatorial in nature, in NQueens there can be multiple solutions
for a given starting board. This allows us to measure the diversity of the generated
solutions. Following GRAM, we track coverage of possible solutions and report the number of
unique solutions generated out of the total number of possible solutions corresponding to
that board. We evaluate on held out initial board states and generate 20 samples per board
state.  (TODO: add more experimental details)

We use the infill layout on an $n\times n$ board (length $n^2$) with token vocabulary
$\{0\!=\!\text{pad},\,1\!=\!\text{empty},\,2\!=\!\text{queen}\}$, sweeping $p_\text{condclean}$ at
two scales: $n=8$ (length $64$) and $n=10$ (length $100$). Training and evaluation splits are
disjoint by clue board. For $n=8$ this yields $1{,}758$ training / $174$ validation puzzles; the
$n=10$ train/validation counts are \textbf{[TODO]}. We train at global batch size $256$ and evaluate
the \emph{last} checkpoint on $750$ held-out puzzles with $20$ generations each
(Table~\ref{tab:nqueens}).

\begin{table}[t]
\centering
\caption{N-Queens infill accuracy and coverage (\%), last checkpoint, $750$ puzzles
$\times\,20$ samples, with $95\%$ bootstrap CIs ($10{,}000$ resamples over input boards) in
brackets. Coverage is monotonic in $p_\text{condclean}$ at both scales and $p{=}0.2$
dominates throughout. Note the accuracy non-monotonicity at $n=10$ ($p{=}0.5 < p{=}1.0$); see text.}
\label{tab:nqueens}
\begin{tabular}{lcccc}
\toprule
& \multicolumn{2}{c}{$n=8$ (length 64)} & \multicolumn{2}{c}{$n=10$ (length 100)} \\
\cmidrule(lr){2-3}\cmidrule(lr){4-5}
$p_\text{condclean}$ & Accuracy & Coverage & Accuracy & Coverage \\
\midrule
$0.2$ & $\mathbf{99.24}$ & $\mathbf{97.31}$ & $\mathbf{94.66}$ & $\mathbf{80.01}$ \\
 & {\scriptsize $[98.86,\,99.55]$} & {\scriptsize $[96.52,\,98.05]$} & {\scriptsize $[94.05,\,95.23]$} & {\scriptsize $[78.73,\,81.35]$} \\
$0.5$ & $98.75$ & $91.41$ & $57.42$ & $71.05$ \\
 & {\scriptsize $[98.20,\,99.23]$} & {\scriptsize $[90.02,\,92.74]$} & {\scriptsize $[55.88,\,58.95]$} & {\scriptsize $[69.31,\,72.77]$} \\
$1.0$ (paste-in) & $97.67$ & $88.07$ & $64.56$ & $64.03$ \\
 & {\scriptsize $[96.73,\,98.51]$} & {\scriptsize $[86.46,\,89.63]$} & {\scriptsize $[62.90,\,66.21]$} & {\scriptsize $[62.37,\,65.70]$} \\
\bottomrule
\end{tabular}
\end{table}

The discriminating effect is in coverage. At $n=8$ accuracy is high throughout ($97.7$--$99.2\%$,
a $\sim1.6$-point spread) while coverage spans $\sim9$ points ($97.31 \to 88.07\%$) and is monotonic
in $p$. At $n=10$, $p=0.2$ dominates decisively on both metrics (a $\sim30$-point accuracy lead and a
$\sim9$-point coverage lead over the next best). Coverage remains monotonic
($80.01 \to 71.05 \to 64.03\%$), but accuracy is \emph{not}: $p=0.5$ ($57.42\%$) falls below
$p=1.0$ ($64.56\%$). The robust cross-task claim is therefore the coverage ordering and the
dominance of heavily-noised conditioning ($p=0.2$), rather than strict accuracy monotonicity. The
$95\%$ bootstrap CIs for the $p{=}0.5$ ($[55.88,58.95]$) and $p{=}1.0$ ($[62.90,66.21]$) accuracies
at $n=10$ do not overlap, so the inversion is not an artifact of evaluation sampling; it is a single
training run, however, so training-seed variance remains unchecked and a multi-seed confirmation is
left to future work.

\section{Analysis and Discussion}
\label{sec:analysis}
\paragraph{Why does noising the conditioning help?}
We offer three (non-exclusive) readings, consistent with the evidence above. First, noising the
clues acts as a \emph{regularizer} against a brittle copy-and-lookup solution: when clues are always
clean, the model can rely on a shortcut $x\to y$ mapping, whereas forcing it to sometimes infer the
answer from noised clues pushes it to internalize the task constraints. Second, this manifests as
\emph{mode coverage}: always-clean training ($p=1$) collapses toward fewer solutions, which is
exactly what the monotonic coverage drop on N-Queens shows (Table~\ref{tab:nqueens}). Third, the
sweep can be read as a \emph{curriculum}: a single $p<1$ run mixes easy (clean-clue) and hard
(noised-clue) conditional prediction problems within training.

As a probe of the first reading, we masked all attention into the non-conditioning tokens at
inference; a model that had merely learned to copy clue $\to$ solution would be largely unaffected,
but solve rate dropped to $0\%$, indicating the model genuinely uses the generated cells rather than
a lookup table.

\paragraph{Relation to classifier-free guidance.}
Because $p$ interpolates between a conditional-only ($p=1$) and unconditional-only ($p=0$) objective,
our scheme can be viewed as a continuous generalization of classifier-free guidance training. We do
not yet apply guidance at sampling time; combining the two is natural future work.

\paragraph{Limitations.}
Our models are small (8-block DDiT) and trained on small datasets, with heavy epoch repetition
(e.g.\ $\sim$1.8k N-Queens-8 examples over $100$k steps at batch $256$), so overfitting/data-efficiency
effects are entangled with the conditioning-noise effect. Our bootstrap CIs control
evaluation-sampling variance, but all runs are single training seeds, so training-seed variance
is unaddressed. Several configuration details remain to be confirmed
(optimizer class, GPU, N-Queens-10 split sizes). We study easy Sudoku and small N-Queens; harder
Sudoku and natural-language reasoning (GSM8k) are in progress and not reported here.

\section{Conclusion}
\label{sec:conclusion}
We revisited a default of conditional continuous diffusion language models---pasting in clean
conditioning tokens at every step---and showed that simply \emph{noising} the conditioning tokens
during training, controlled by a single probability $p_\text{condclean}$, consistently improves
conditional generation. Lower $p$ (more noising) raises Sudoku infill accuracy and, most clearly,
N-Queens solution coverage, with an intermediate-but-low $p\approx0.2$ best across tasks. The method
is a one-line change to the training corruption and can be seen as a continuous generalization of
classifier-free guidance. Future work includes the random-conditioning-time variant, larger models
and harder Sudoku, sampling-time guidance, multi-seed variance checks, and language reasoning (GSM8k).

\bibliography{references}
\bibliographystyle{colm2026_conference}

\appendix
\section{Full Hyperparameters}
\label{app:hyperparams}
Table~\ref{tab:hparams} lists the shared training configuration (N-Queens Run; Sudoku differs only
in board length $81$, batch size $32$, and $160$k training steps).

\begin{table}[h]
\centering
\caption{Shared hyperparameters. \textbf{[TODO]} items remain to be confirmed against the original
configuration.}
\label{tab:hparams}
\begin{tabular}{lll}
\toprule
Group & Setting & Value \\
\midrule
Model & hidden size / blocks / heads & $512$ / $8$ / $8$ \\
      & cond.\ dim / dropout & $128$ / $0.1$ \\
      & board length & $81$ (Sudoku), $64$/$100$ (N-Queens $n{=}8$/$10$) \\
Algorithm & parameterization / pred.\ type & mean / $x_1$ \\
          & loss type & flow \\
          & diffusion forcing & true \\
          & $p_\text{condclean}$ & $\{0.2,0.5,1.0\}$ (swept) \\
Noise & type / eps & log-linear / $0$ \\
Training & max steps & $100{,}001$ (N-Queens), $160$k (Sudoku) \\
         & global batch size & $256$ (N-Queens), $32$ (Sudoku) \\
         & precision / EMA & bf16-mixed / $0.9999$ \\
         & grad clip & $1.0$ \\
         & $t$-curriculum & $0.0\to1.0$ over $100$k steps \\
Optimizer & class & Adam/AdamW \textbf{[TODO]} \\
          & lr / $(\beta_1,\beta_2)$ / wd & $3\!\times\!10^{-4}$ / $(0.9,0.999)$ / $0$ \\
          & warmup & $2{,}500$ steps (constant after) \\
Sampling & predictor / solver & ancestral / Euler \\
         & steps / temperature / nucleus & $128$ / $1.0$ / $1.0$ \\
Hardware & accelerator / GPU & cuda, 1 device / \textbf{[TODO]} \\
\bottomrule
\end{tabular}
\end{table}

\end{document}
